The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-28 paper.tex
Revisao da Secao 2.2.

Embedded applications make extensive use of shared memory, such as for sharing 
resources through memory-mapped registers.
Hence, an important goal of \CEU is to ensure a reliable execution for 
concurrent programs that share memory.
%
Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{"call 
f1() || call f2()"}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
Therefore, Esterel forbids sharing memory: \emph{``if a variable is written by 
some thread, then it can neither be read nor be written by concurrent 
threads''}~\cite{esterel.primer}.

@@

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows~\cite{ceu.sensys13}:
\emph{if a variable is written in a trail segment, then a concurrent trail 
segment cannot read or write to that variable, nor dereference a pointer of 
that variable type.}
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).
The algorithm for the analysis holds the set of all events in preceding 
\code{await} statements for each variable access.
Then, the sets for all accesses in parallel trails are compared to assert that 
no events are shared among them.
Otherwise the compiler warns about the suspicious accesses.

Regardless of the static analysis, when multiple trails are active at a time in 
\CEU, they are scheduled in the order they appear in the program source code.
%
Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
suspicious, the assignments to \code{y} both atomic and deterministic, i.e., 
after \code{A} occurs, \code{y} is $4$ (\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
[
    call f1();
$\|$
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
par/and do
    _f1();
with
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2}.
\label{lst.det}
}
\end{figure}

>>>>

Embedded applications make extensive use of shared memory, such as for sharing 
resources through memory-mapped registers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
with concurrent lines of execution sharing memory.
%
Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{<<call 
f1() || call f2()>>}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
Therefore, Esterel forbids sharing memory between lines of execution: 
\emph{``if a variable is written by some thread, then it can neither be read 
nor be written by concurrent threads''}~\cite{esterel.primer}.

@@

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows~\cite{ceu.sensys13}:
\emph{if a variable is written in a trail segment, then a concurrent trail 
segment cannot read or write to that variable, nor dereference a pointer of 
that variable type.}
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).
%The algorithm for the analysis holds the set of all events in preceding 
%\code{await} statements for each variable access.
%Then, the sets for all accesses in parallel trails are compared to assert that 
%no events are shared among them.
%Otherwise the compiler warns about the suspicious accesses.

Regardless of the temporal analysis of \CEU, when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.
%
Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
suspicious, the assignments to \code{y} are both atomic and deterministic, 
i.e., after the reaction to \code{A} terminates, the value of \code{y} is $4$ 
(\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;
[
    await A;
    call f1();
$\|$
    await A;
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// CEU
input void A;
par/and do
    await A;
    _f1();
with
    await A;
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2} due to 
deterministic scheduling based on lexical order.
\label{lst.det}
}
\end{figure}
